---
title: "On harmonic oscillator hazard functions"
author: '[F. Javier Rubio](https://sites.google.com/site/fjavierrubio67/) and J. Andres Christen'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
bibliography: references.bib
---

# Harmonic oscillator hazard function

This document provides codes and information about the *Damped Harmonic Oscillator* model proposed in @christen:2024. 

## The damped harmonic oscillator model

The hazard-Response model is defined through the second order ODE:
\begin{equation}
h''(t) + 2 \eta w_0 h'(t) + w^2_0 (h(t) - h_b) = 0,
\label{eq:harmonic_osc}    
\end{equation}
where $w_0 > 0$ is called the ``natural frequency'', $\eta >0$ is the ``damping coefficient''. The shift parameter $h_b > 0$ ensures that the solution is positive


The following R code presents an example from @christen:2024 where we fit the damped harmonic oscillator model to the `rotterdam` data set from the R package `survival`. We present both Likelihood-based and Bayesian analyses.


# Data preparation
```{r message=FALSE}
#######################################################################################
# Data preparation
#######################################################################################

rm(list=ls())

# Required packages
library(deSolve)
library(survival)
library(ggplot2)
#library(devtools)
#install_github("FJRubio67/HazReg")
library(HazReg)
library(spBayes)
library(Rtwalk)
library(knitr)
library(bshazard)
```

```{r}
source("routinesHO.R")
```

```{r}
#--------------------------------------------------------------------------------------------------
#--------------------------------------------------------------------------------------------------
# Data preparation
#--------------------------------------------------------------------------------------------------
#--------------------------------------------------------------------------------------------------


head(rotterdam)

dim(rotterdam)

# Kaplan-Meier estimator for the survival times
km <- survfit(Surv(rotterdam$dtime/365.24, rotterdam$death) ~ 1)

plot(km$time, km$surv, type = "l", col = "black", lwd = 2, lty = 1, 
     ylim = c(0,1), xlab = "Time", ylab = "Survival")



# New data frame: logical status, time in years, survival times sorted
df <- data.frame(time = rotterdam$dtime, status = rotterdam$death)
df$status <- as.logical(rotterdam$death)
df$time <- df$time/365.24

df <- df[order(df$time),]

# Required quantities
status <- as.logical(df$status)
t_obs <- df$time[status]
survtimes <- df$time
```

# Maximum Likelihood Analysis

## Weibull Model

```{r}
#==================================================================================================
# Maximum Likelihood Analysis
#==================================================================================================


#--------------------------------------------------------------------------------------------------
# Fitting a Weibull distribution
#--------------------------------------------------------------------------------------------------

# Initial value
initW <- c(0,0)

# Optimisation step
OPTW <- GHMLE(initW, survtimes, status, hstr = "baseline", dist = "Weibull", method = "nlminb", maxit = 10000)

MLEW <- exp(OPTW$OPT$par)

# Fitted Weibull hazard
fithw <- Vectorize(function(t) hweibull( t, exp(OPTW$OPT$par[1]), exp(OPTW$OPT$par[2]) ) )

# Fitted Weibull cumulative hazard
fitchw <- Vectorize(function(t) chweibull( t, exp(OPTW$OPT$par[1]), exp(OPTW$OPT$par[2]) ) )

# Fitted Weibull survival
fitsw <- Vectorize(function(t) exp(-chweibull( t, exp(OPTW$OPT$par[1]), exp(OPTW$OPT$par[2]) ) ))

# AIC
AICW <- 2*OPTW$OPT$objective + 2*length(OPTW$OPT$par)

# BIC
BICW <- 2*OPTW$OPT$objective + length(OPTW$OPT$par)*log(length(survtimes))
```

## The Power Generalised Weibull Model

```{r}
#==================================================================================================
# Maximum Likelihood Analysis
#==================================================================================================

#--------------------------------------------------------------------------------------------------
# Fitting a PGW distribution
#--------------------------------------------------------------------------------------------------

# Initial value
initPGW <- c(0,0,0)

# Optimisation step
OPTPGW <- GHMLE(initPGW, survtimes, status, hstr = "baseline", dist = "PGW", 
                method = "nlminb", maxit = 10000)

MLEPGW <- exp(OPTPGW$OPT$par)

# Fitted Weibull hazard
fithpgw <- Vectorize(function(t) hpgw( t, MLEPGW[1], MLEPGW[2], MLEPGW[3] ) )

# Fitted Weibull cumulative hazard
fitchpgw <- Vectorize(function(t) chpgw( t, MLEPGW[1], MLEPGW[2], MLEPGW[3] ) )

# Fitted Weibull survival
fitspgw <- Vectorize(function(t) exp(-chpgw( t, MLEPGW[1], MLEPGW[2], MLEPGW[3] ) ))

# AIC
AICPGW <- 2*OPTPGW$OPT$objective + 2*length(OPTPGW$OPT$par)

# BIC
BICPGW <- 2*OPTPGW$OPT$objective + length(OPTPGW$OPT$par)*log(length(survtimes))

```

## The Damped Harmonic Oscillator Model

```{r}
#==================================================================================================
# MLE Analysis
#==================================================================================================

#--------------------------------------------------------------------------------------------------
# Harmonic Oscillator model for the hazard function: Solver solution
#--------------------------------------------------------------------------------------------------

# Initial conditions

# Survival at time 0
S0 <- 1
# Survival at 1 month
St <- 0.999
# Survival at 2 months
Stt <- 0.998
# dt = 1 month
dt <- 1/12

# Approximation of the initial conditions
Sp <- -(St-S0)/dt
Spp <- (Stt - 2*St + S0)/(dt^2)

h00 <- Sp/St
r00 <- h00^2 - Spp/St

# Initial point
initHO <- c(0.1,0,0)

# Optimisation step
OPTHO <- optim(initHO, log_likHO, control = list(maxit = 1000))
OPTHO

MLEHO <- exp(OPTHO$par)

# AIC
AICHO <- 2*OPTHO$value + 2*length(OPTHO$par)

# BIC
BICHO <- 2*OPTHO$value + length(OPTHO$par)*log(length(survtimes))
```

## Comparison

```{r}
# AIC comparison
AICs <- c(AICW, AICPGW, AICHO)

AICs

# Best model (Harmonic oscillator)
which.min(AICs)

# BIC comparison
BICs <- c(BICW, BICPGW, BICHO)

BICs

# Best model (Harmonic oscillator)
which.min(BICs)

# Fitted hazard, cumulative hazard, and survival
fithHO <- Vectorize(function(t) hHO(t,exp(OPTHO$par[1]), exp(OPTHO$par[2]), exp(OPTHO$par[3]), h00, r00))
fitchHO <- Vectorize(function(t) chHO(t,exp(OPTHO$par[1]), exp(OPTHO$par[2]), exp(OPTHO$par[3]), h00, r00))
fitsHO <- Vectorize(function(t) exp(-chHO(t,exp(OPTHO$par[1]), exp(OPTHO$par[2]), exp(OPTHO$par[3]), h00, r00)))

# Comparison: hazard functions
curve(fithHO, 0, max(survtimes), xlim= c(0,max(survtimes)), ylim = c(0,0.125), lwd = 2,
     xlab = "Time", ylab = "Hazard", main = "Harmonic Oscillator", cex.axis = 1.5, cex.lab = 1.5)
curve(fithw, 0, max(survtimes), lwd= 2, lty = 2, col = "gray", add = TRUE)
curve(fithpgw, 0, max(survtimes), lwd= 2, lty = 3, col = "gray", add = TRUE)
legend("topleft", legend = c("Harmonic Oscillator","Weibull", "PGW"), lty = c(1,2,3), 
       lwd = c(2,2,2), col = c("black","gray","gray"))

# Comparison: cumulative hazard functions
curve(fitchHO, 0, max(survtimes), xlim = c(0,max(survtimes)), ylim = c(0,2.25),  lwd = 2,
     xlab = "Time", ylab = "Cumulative Hazard", main = "Harmonic Oscillator", cex.axis = 1.5, cex.lab = 1.5)
curve(fitchw, 0, max(survtimes), lwd= 2, lty = 2, col = "gray", add = TRUE)
curve(fitchpgw, 0, max(survtimes), lwd= 2, lty = 3, col = "gray", add = TRUE)
legend("topleft", legend = c("Harmonic Oscillator","Weibull","PGW"), lty = c(1,2,3), 
       lwd = c(2,2,2), col = c("black","gray","gray"))

# Comparison: survival functions
curve(fitsHO, 0, max(survtimes), xlim = c(0,max(survtimes)), ylim = c(0,1), lwd = 2,
     xlab = "Time", ylab = "Survival", main = "Harmonic Oscillator", cex.axis = 1.5, cex.lab = 1.5)
curve(fitsw, 0, max(survtimes), lwd= 2, lty = 2, col = "gray", add = TRUE)
curve(fitspgw, 0, max(survtimes), lwd= 2, lty = 3, col = "gray", add = TRUE)
points(km$time, km$surv, type = "l", col = "green", lwd = 2, lty = 1)
legend("topright", legend = c("Harmonic Oscillator","Weibull","PGW","KM"), lty = c(1,2,3,1), 
       lwd = c(2,2,2,2), col = c("black","gray","gray","green"))

```

# Bayesian Analysis


## Weibull
```{r}
#==================================================================================================
# Bayesian Analysis
#==================================================================================================

#--------------------------------------------------------------------------------------------------
# Priors
#--------------------------------------------------------------------------------------------------

par(mfrow = c(1,2))
p_sigmaW <- Vectorize(function(t) dgamma(t, shape = 2, scale = 2))
curve(p_sigmaW,0,15, n = 1000, xlab = ~sigma, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2)

p_nuW <- Vectorize(function(t) dgamma(t, shape = 2, scale = 2))
curve(p_nuW,0,15, n = 1000, xlab = ~nu, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2)


par(mfrow = c(1,1))

#--------------------------------------------------------------------------------------------------
# Weibull model for the hazard function
#--------------------------------------------------------------------------------------------------

# Support
SupportW <- function(x) {   TRUE }

# Random initial points
X0W <- function(x) { OPTW$OPT$par + runif(2,-0.01,0.01) }

# twalk for analytic solution
set.seed(1234)
infoW <- Runtwalk( dim=2,  Tr=110000,  Obj=log_postW, Supp=SupportW, 
                   x0=X0W(), xp0=X0W(), PlotLogPost = FALSE) 


# Posterior sample after burn-in and thinning
ind=seq(10000,110000,100) 

# Summaries
summW <- apply(exp(infoW$output[ind,]),2,summary)
colnames(summW) <- c("sigma","nu")
kable(summW, digits = 3)

# KDEs
sigmapW <- exp(infoW$output[,1][ind])
nupW <- exp(infoW$output[,2][ind])

plot(density(sigmapW), main = "", xlab = expression(sigma), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_sigmaW,13,17, n = 1000, xlab = ~lambda, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)


plot(density(nupW), main = "", xlab = expression(nu), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_nuW,1.1,1.5, n = 1000, xlab = ~kappa, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)
```

## The Power Generalised Weibull Model
```{r}
#==================================================================================================
# Bayesian Analysis
#==================================================================================================

#--------------------------------------------------------------------------------------------------
# Priors
#--------------------------------------------------------------------------------------------------

par(mfrow = c(1,3))
p_sigmaPGW <- Vectorize(function(t) dgamma(t, shape = 2, scale = 2))
curve(p_sigmaPGW,0,15, n = 1000, xlab = ~sigma, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2)

p_nuPGW <- Vectorize(function(t) dgamma(t, shape = 2, scale = 2))
curve(p_nuPGW,0,15, n = 1000, xlab = ~nu, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2)

p_gammaPGW <- Vectorize(function(t) dgamma(t, shape = 2, scale = 2))
curve(p_gammaPGW,0,15, n = 1000, xlab = ~gamma, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2)

par(mfrow = c(1,1))

#--------------------------------------------------------------------------------------------------
# Weibull model for the hazard function
#--------------------------------------------------------------------------------------------------

# Support
SupportPGW <- function(x) {   TRUE }

# Random initial points
X0PGW <- function(x) { OPTPGW$OPT$par + runif(3,-0.01,0.01) }


# twalk for analytic solution
set.seed(1234)
infoPGW <- Runtwalk( dim=3,  Tr=110000,  Obj=log_postPGW, Supp=SupportPGW, 
                     x0=X0PGW(), xp0=X0PGW(), PlotLogPost = FALSE) 


# Posterior sample after burn-in and thinning
ind=seq(10000,110000,100) 

# Summaries
summPGW <- apply(exp(infoPGW$output[ind,]),2,summary)
colnames(summPGW) <- c("sigma","nu","gamma")
kable(summPGW, digits = 3)

# KDEs
sigmapPGW <- exp(infoPGW$output[,1][ind])
nupPGW <- exp(infoPGW$output[,2][ind])
gammapPGW <- exp(infoPGW$output[,3][ind])

plot(density(sigmapPGW), main = "", xlab = expression(sigma), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_sigmaPGW,2,5, n = 1000, xlab = ~lambda, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)


plot(density(nupPGW), main = "", xlab = expression(nu), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_nuPGW,1.5,2.7, n = 1000, xlab = ~kappa, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)

plot(density(gammapPGW), main = "", xlab = expression(gamma), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_gammaPGW,0,15, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)
```

## The Hazard-Response Model

```{r}
#==================================================================================================
# Bayesian Analysis
#==================================================================================================

#--------------------------------------------------------------------------------------------------
# Priors
#--------------------------------------------------------------------------------------------------

par(mfrow = c(2,2))
p_eta <- Vectorize(function(t) dgamma(t, shape = 0.5, scale = 1))
curve(p_eta,0,2, n = 1000, xlab = ~eta, ylab = "Prior Density",
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2)

p_w0 <- Vectorize(function(t) dgamma(t, shape = 2, scale = 1))
curve(p_w0,0,10, n = 1000, xlab = expression(w_0), ylab = "Prior Density",
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2)

p_hb <- Vectorize(function(t) dgamma(t, shape = 0.1, scale = 0.1))
curve(p_hb,0,0.1, n = 1000, xlab = expression(h_b), ylab = "Prior Density",
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2)


par(mfrow = c(1,1))


#--------------------------------------------------------------------------------------------------
# Hazard-Response ODE model for the hazard function: Solver solution
#--------------------------------------------------------------------------------------------------


n.batch <- 1100
batch.length <- 100

lp <- function(par) -log_postHO(par)

inits <- OPTHO$par
```

```{r include=FALSE}
set.seed(1234)
infoHO<- adaptMetropGibbs(ltd=lp, starting=inits, accept.rate=0.44, batch=n.batch, 
                           batch.length=batch.length, report=100, verbose=FALSE)
```

```{r eval=FALSE}
set.seed(1234)
infoHO <- adaptMetropGibbs(ltd=lp, starting=inits, accept.rate=0.44, batch=n.batch, 
                           batch.length=batch.length, report=100, verbose=FALSE)
```

```{r}
chainHO <- infoHO$p.theta.samples[,1:3]

# Burning and thinning the chain
burn <- 1e4
thin <- 100
NS <- n.batch*batch.length
ind <- seq(burn,NS,thin)



# Summaries
summHR <- apply(exp(chainHO[ind,1:3]),2,summary)
colnames(summHR) <- c("eta","w_0","h_b")
kable(summHR, digits = 3)

# KDEs
etaHO <- exp(chainHO[,1][ind])
w0HO <- exp(chainHO[,2][ind])
hbHO <- exp(chainHO[,3][ind])


plot(density(etaHO), main = "", xlab = expression(lambda), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_eta,0,1, n = 1000, xlab = ~lambda, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)


plot(density(w0HO), main = "", xlab = expression(kappa), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_w0,0.5,2, n = 1000, xlab = expression(w_0), ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)

plot(density(hbHO), main = "", xlab = expression(alpha), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_hb,0,0.1, n = 1000, xlab = expression(h_b), ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)

```


## Predictive Hazard functions 

```{r} 
#---------------------------------------------------------------------------------------
# Predictive Hazard functions
#---------------------------------------------------------------------------------------

# Predictive Weibull hazard
predhW <- Vectorize(function(t){
  num <- den <- temp <- vector()
  for(i in 1:length(ind)){
    num[i] <- exp(-chweibull( t, sigmapW[i], nupW[i]))*hweibull( t, sigmapW[i], nupW[i])
    den[i] <- exp(-chweibull( t, sigmapW[i], nupW[i]))
  }
  return(mean(num)/mean(den))
})

# Predictive PGW
predhPGW <- Vectorize(function(t){
  num <- den <- temp <- vector()
  for(i in 1:length(ind)) num[i] <- exp(-chpgw( t, sigmapPGW[i], nupPGW[i], gammapPGW[i]))*
      hpgw( t, sigmapPGW[i], nupPGW[i], gammapPGW[i])
  for(i in 1:length(ind)) den[i] <- exp(-chpgw( t, sigmapPGW[i], nupPGW[i], gammapPGW[i]))
  return(mean(num)/mean(den))
})



# Creating the credible envelopes
tvec <- seq(0,20,by = 0.01)
ntvec <- length(tvec)

# Weibull
hCIW <- matrix(0, ncol = ntvec, nrow = length(ind))

for(j in 1:length(ind)){
  for(k in 1:ntvec){
    hCIW[j,k ] <- hweibull( tvec[k],sigmapW[j], nupW[j]) 
  }
} 

hW <-  predhW(tvec)


hCIWL <- apply(hCIW, 2, ql)
hCIWU <- apply(hCIW, 2, qu)

# PGW
hCIPGW <- matrix(0, ncol = ntvec, nrow = length(ind))

for(j in 1:length(ind)){
  for(k in 1:ntvec){
    hCIPGW[j,k ] <- hpgw( tvec[k],sigmapPGW[j], nupPGW[j], gammapPGW[j]) 
  }
} 

hPGWs <-  predhPGW(tvec)

hCIPGWL <- apply(hCIPGW, 2, ql)
hCIPGWU <- apply(hCIPGW, 2, qu)

# Hazard-Response

hCIHR <- matrix(0, ncol = ntvec, nrow = length(ind))
chCIHR <- matrix(0, ncol = ntvec, nrow = length(ind))
SCIHR <- matrix(0, ncol = ntvec, nrow = length(ind))
qCIHR <- matrix(0, ncol = ntvec, nrow = length(ind))

for(j in 1:length(ind)){
  paramsHRj  <- c(lambda = lambdapHR[j], kappa = kappapHR[j], alpha = alphapHR[j],
                  beta = betapHR[j])
  initHRj      <- c(h = h0, q = q0, H = 0 )
  outj <- ode(initHRj, tvec, hazmodHR, paramsHRj, method = "lsode", jacfunc = jacODE, jactype = "fullusr")
  
  hCIHR[j, ] <- outj[,2]
  qCIHR[j, ] <- outj[,3]
  chCIHR[j, ] <- outj[,4]
  SCIHR[j, ] <- exp(-outj[,4])
  
} 


numHR <- hCIHR*exp(-chCIHR)
denHR <- exp(-chCIHR)

hpredHR <- colMeans(numHR)/colMeans(denHR)


hCIHRL <- apply(hCIHR, 2, ql)
hCIHRU <- apply(hCIHR, 2, qu)

qCIHRL <- apply(qCIHR, 2, ql)
qCIHRU <- apply(qCIHR, 2, qu)


# Plots

# Weibull
plot(tvec,  hW, type = "l", ylim = c(0,0.15), xlab = "Time", ylab = "Predictive Hazard", 
     cex.axis = 1.5, cex.lab = 1.5, lwd =1)
points(tvec,  hCIWL, col = "gray", type = "l")
points(tvec,  hCIWU, col = "gray", type = "l")
polygon(c(tvec, rev(tvec)), c(hCIWL[order(tvec)], rev(hCIWU[order(tvec)])),
        col = "gray", border = NA)
points(tvec,  hW,type = "l", col = "black", lwd = 1)

# PGW
points(tvec,  hPGWs, type = "l", xlab = "Time", ylab = "Predictive Hazard", 
       cex.axis = 1.5, cex.lab = 1.5, lwd =1, lty = 2)
points(tvec,  hCIPGWL, col = "gray", type = "l")
points(tvec,  hCIPGWU, col = "gray", type = "l")
polygon(c(tvec, rev(tvec)), c(hCIPGWL[order(tvec)], rev(hCIPGWU[order(tvec)])),
        col = "gray", border = NA)
points(tvec,  hPGWs,type = "l", col = "black", lwd = 1, lty =2)

# Hazard-Response
points(tvec,  hpredHR, type = "l", xlab = "Time", ylab = "Predictive Hazard", 
       cex.axis = 1.5, cex.lab = 1.5, lwd =1, lty = 3)
points(tvec,  hCIHRL, col = "gray", type = "l")
points(tvec,  hCIHRU, col = "gray", type = "l")
polygon(c(tvec, rev(tvec)), c(hCIHRL[order(tvec)], rev(hCIHRU[order(tvec)])),
        col = "gray", border = NA)
points(tvec,  hpredHR,type = "l", col = "black", lwd = 1, lty =3)
points(tvec,  hPGWs,type = "l", col = "black", lwd = 1, lty =2)
legend("bottomright", legend = c("Weibull", "PGW", "HR"), lty = c(1,2,3), 
       lwd = c(2,2,2), col = c("black","black","black"))



``` 


## Predictive Hazard vs. Response

```{r}
# Hazard vs. Response
plot(tvec,  hpredHR, type = "l", xlab = "Time", ylab = "Predictive Hazard", 
       cex.axis = 1.5, cex.lab = 1.5, lwd =2, lty = 1, ylim = c(0,0.1))

qpredHR <- colMeans(qCIHR)

points(tvec,  qpredHR, type = "l", xlab = "Time", ylab = "Predictive Hazard", 
       cex.axis = 1.5, cex.lab = 1.5, lwd =2, lty = 2)

legend("bottomright", legend = c("Hazard","Response"), lty = c(1,2), 
       lwd = c(2,2), col = c("black","black"))
```

## Predictive Survival functions 

```{r}
#--------------------------------------------------------------------------------------- 
# Predictive Survival functions 
#--------------------------------------------------------------------------------------- 

# Predictive Weibull survival 
predsW <- Vectorize(function(t){ 
  temp <- vector() 
  for(i in 1:length(ind)) temp[i] <- exp(-chweibull(t,sigmapW[i], nupW[i]) )  
  return(mean(temp)) 
}) 

# Predictive PGW survival 
predsPGW <- Vectorize(function(t){ 
  temp <- vector() 
  for(i in 1:length(ind)) temp[i] <- exp(-chpgw( t,sigmapPGW[i], nupPGW[i], gammapPGW[i]) )  
  return(mean(temp)) 
}) 


# Predictive Hazard-Response survival: Solver solution 
predsHR <- colMeans(denHR)


# Weibull 
SCIW <- matrix(0, ncol = ntvec, nrow = length(ind)) 

for(j in 1:length(ind)){ 
  for(k in 1:ntvec){ 
    SCIW[j,k ] <- exp(-chweibull( tvec[k],sigmapW[j], nupW[j])) 
  } 
}

SW <-  predsW(tvec) 

SCIWL <- apply(SCIW, 2, ql) 
SCIWU <- apply(SCIW, 2, qu) 

# PGW

SCIPGW <- matrix(0, ncol = ntvec, nrow = length(ind)) 

for(j in 1:length(ind)){ 
  for(k in 1:ntvec){ 
    SCIPGW[j,k ] <- exp(-chpgw( tvec[k],sigmapPGW[j], nupPGW[j], gammapPGW[j])) 
  } 
}

SPGW <-  predsPGW(tvec) 

SCIPGWL <- apply(SCIPGW, 2, ql) 
SCIPGWU <- apply(SCIPGW, 2, qu) 


# Hazard-Response 
SpredCIHR <- colMeans(denHR) 

SCIHRL <- apply(SCIHR, 2, ql) 
SCIHRU <- apply(SCIHR, 2, qu) 

# Plots (no credible envelopes) 

# Weibull 
plot(tvec,  SW, type = "l", ylim = c(0,1), xlab = "Time", ylab = "Predictive Survival",  
     cex.axis = 1.5, cex.lab = 1.5, lwd =1) 

# PGW 
points(tvec,  SPGW, type = "l", ylim = c(0,1), xlab = "Time", ylab = "Predictive Survival",  
       cex.axis = 1.5, cex.lab = 1.5, lwd =1, lty = 2) 

# Hazard-Response 
points(tvec,  SpredCIHR, type = "l", ylim = c(0,1), xlab = "Time", ylab = "Predictive Survival",  
       cex.axis = 1.5, cex.lab = 1.5, lwd =1, lty = 3) 

legend("topright", legend = c("Weibull", "PGW", "HR"), lty = c(1,2,3),  
       lwd = c(2,2,2), col = c("black","black","black")) 


# Plots (with credible envelopes) 

# Weibull 
plot(tvec,  SW, type = "l", ylim = c(0,1), xlab = "Time", ylab = "Predictive Survival",  
     cex.axis = 1.5, cex.lab = 1.5, lwd =1) 
points(tvec,  SCIWL, col = "gray", type = "l") 
points(tvec,  SCIWU, col = "gray", type = "l") 
polygon(c(tvec, rev(tvec)), c(SCIWL[order(tvec)], rev(SCIWU[order(tvec)])), 
        col = "gray", border = NA) 
points(tvec,  SW,type = "l", col = "black", lwd = 1) 

# PGW 
points(tvec,  SPGW, type = "l", ylim = c(0,1), xlab = "Time", ylab = "Predictive Survival",  
       cex.axis = 1.5, cex.lab = 1.5, lwd =1, lty = 2) 
points(tvec,  SCIPGWL, col = "gray", type = "l") 
points(tvec,  SCIPGWU, col = "gray", type = "l") 
polygon(c(tvec, rev(tvec)), c(SCIPGWL[order(tvec)], rev(SCIPGWU[order(tvec)])), 
        col = "gray", border = NA) 
points(tvec,  SPGW,type = "l", col = "black", lwd = 1, lty =2) 

# Hazard-Response 
points(tvec,  SpredCIHR, type = "l", ylim = c(0,1), xlab = "Time", ylab = "Predictive Survival",  
       cex.axis = 1.5, cex.lab = 1.5, lwd =1, lty = 3) 
points(tvec,  SCIHRL, col = "gray", type = "l") 
points(tvec,  SCIHRU, col = "gray", type = "l") 
polygon(c(tvec, rev(tvec)), c(SCIHRL[order(tvec)], rev(SCIHRU[order(tvec)])), 
        col = "gray", border = NA) 
points(tvec,  SpredCIHR,type = "l", col = "black", lwd = 1, lty =3) 

points(tvec,  SPGW,type = "l", col = "black", lwd = 1, lty =2) 

legend("topright", legend = c("Weibull", "PGW", "HR"), lty = c(1,2,3),  
       lwd = c(2,2,2), col = c("black","black","black")) 

```


# Comparison with a spline-based estimator
Now, we present a comparison of the estimated hazard function with that obtained with the R command `bshazard` from the R package `bshazard`. This estimator is based on B-Splines. See [bshazard: Nonparametric Smoothing of the Hazard Function](https://cran.r-project.org/package=bshazard).

```{r message=FALSE}
# fitting the estimator
fit <- bshazard(Surv(df$time, df$status) ~ 1, data = df, nbin = 100, degree = 3, verbose = FALSE)
```

```{r}
# Hazard vs. Response
plot(tvec,  hpredHR, type = "l", xlab = "Time", ylab = "Hazard", 
       cex.axis = 1.5, cex.lab = 1.5, lwd =2, lty = 1, ylim = c(0,0.125))

# bshazard with confidence intervals
points(fit$time, fit$hazard, type='l', lwd = 2, ylim = c(0,0.15), col = "darkgray")
lines(fit$time, fit$lower.ci, lty = 2, lwd = 1, col = "gray")
lines(fit$time, fit$upper.ci, lty = 2, lwd = 1, col = "gray")

# Hazard vs. Response
points(tvec,  hpredHR, type = "l", xlab = "Time", ylab = "Hazard", 
     cex.axis = 1.5, cex.lab = 1.5, lwd =2, lty = 1, ylim = c(0,0.15))

```

# Sensitivity Analysis

In this section, we provide a sensitivity analysis where instead of fixing the initial conditions $h_0$ and $q_0$, we set a prior on these parameters. These priors are consistent with the choice of the initial conditions chosen in the previous study, in the sense that their mean coincides with the values previously chosen. For $h_0$ we choose a gamma prior with shape parameter $2$ and scale $5\times 10^{-3}$. For $q_0$ we choose a gamma prior with shape parameter $2$ and scale $5\times 10^{-7}$. We can see that the posterior distribution of $h_0$ and $q_0$ is very similar to the prior, indicating weak identifiability of these parameters. 


```{r}

# Log-posterior (unknown initial conditions)
lpX <- function(par) -log_postHRX(par)

initsX <- c(OPTHR$par,h0,q0)


## ----include=FALSE----------------------------------------------------------------------------------------
set.seed(1234)
infoHRX <- adaptMetropGibbs(ltd=lpX, starting=initsX, accept.rate=0.44, batch=n.batch, 
                           batch.length=batch.length, report=100, verbose=FALSE)


chainHRX <- infoHRX$p.theta.samples[,1:6]


# Summaries
summHRX <- apply(exp(chainHRX[ind,1:6]),2,summary)
colnames(summHRX) <- c("lambda","kappa","alpha","beta","h0","q0")
kable(summHRX, digits = 3)

# KDEs
lambdapHRX <- exp(chainHRX[,1][ind])
kappapHRX <- exp(chainHRX[,2][ind])
alphapHRX <- exp(chainHRX[,3][ind])
betapHRX <- exp(chainHRX[,4][ind])
h0pHRX <- exp(chainHRX[,5][ind])
q0pHRX <- exp(chainHRX[,6][ind])


plot(density(lambdapHRX), main = "", xlab = expression(lambda), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_lambdaHR,0,3, n = 1000, xlab = ~lambda, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)


plot(density(kappapHRX), main = "", xlab = expression(kappa), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_kappaHR,0.05,0.3, n = 1000, xlab = ~kappa, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)

plot(density(alphapHRX), main = "", xlab = expression(alpha), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_alphaHR,0,10, n = 1000, xlab = ~alpha, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)

plot(density(betapHRX), main = "", xlab = expression(beta), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_betaHR,0,15, n = 1000, xlab = ~beta, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)

p_h0HR <- Vectorize(function(t) dgamma(t, shape = 2, scale = 5e-3))
plot(density(h0pHRX), main = "", xlab = expression(h[0]), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_h0HR,0,0.05, n = 1000, xlab = ~beta, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)

p_q0HR <- Vectorize(function(t) dgamma(t, shape = 2, scale = 5e-7))
plot(density(q0pHRX), main = "", xlab = expression(h[0]), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
curve(p_q0HR,0,7e-6, n = 1000, xlab = ~beta, ylab = "Prior Density", 
      cex.axis = 1.5, cex.lab = 1.5, lwd = 2, lty = 2, add = TRUE)


# Comparison with the posteriors obtained by fixing the initial conditions

plot(density(lambdapHR), main = "", xlab = expression(lambda), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
points(density(lambdapHRX), lwd = 2, type = "l", lty = 2)
legend("topright", legend = c("fixed","prior"), lty = c(1,2), lwd = c(2,2))


plot(density(kappapHR), main = "", xlab = expression(kappa), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2, ylim = c(0,30))
points(density(kappapHRX), lwd = 2, type = "l", lty = 2)
legend("topright", legend = c("fixed","prior"), lty = c(1,2), lwd = c(2,2))


plot(density(alphapHR), main = "", xlab = expression(alpha), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
points(density(alphapHRX), lwd = 2, type = "l", lty = 2)
legend("topright", legend = c("fixed","prior"), lty = c(1,2), lwd = c(2,2))

plot(density(betapHR), main = "", xlab = expression(beta), ylab = "Density",
     cex.axis = 1.5, cex.lab = 1.5, lwd = 2)
points(density(betapHRX), lwd = 2, type = "l", lty = 2)
legend("topright", legend = c("fixed","prior"), lty = c(1,2), lwd = c(2,2))

```